---
title: "Homework2: Exploring Gapminder dataset using dplyr"
author: "Jummy David"
date: "September 25, 2018"
output:
  github_document:
    toc: yes

---

In this homework, we will explore different variables. To start with, load the `tidyverse` and `gapminder` R packages. 

```{r}
library(gapminder)
library(tidyverse)
?gapminder # to know more about the dataset
```

## Smell testing the data:

Looking at the class of a dataset, is it a data.frame? `class` can be used to check if the dataset is a data frame.

```{r}
head(gapminder)
class(gapminder)
```


`Structure` which can be written as `str` is used as a single command to smell check the data. It also shows that it is a data frame, with 1704 observations and 6 variables. It gives other information about the data. 

```{r}
str(gapminder)
```

Checking the number of rows and columns with `nrow` and `ncol` respectively and as well with `dim`. 

```{r}
nrow(gapminder) #checks the number of rows/observations (1704)
ncol(gapminder) #checks the number of columns/variables (6)
dim(gapminder) #the number of rows/observations (1704) and columns/variables (6)
```

Extent/size of a dataset

```{r}
length(gapminder) #Shows that it is of 6 variables
dim(gapminder) #the number of rows/observations (1704) and columns/variables (6)

```
In summary, we have;

Data type in each variable. It gives each variable name and data type.


|     Variables         | Data type                        |
|-----------------------|----------------------------------|
| country               | `categorical/qualitative/factor` |
| continent             | `categorical/qualitative/factor` |
| year                  |  `quantitative/integer` |
| lifeExp               |  `quantitative/numeric` |
| pop                   | `quantitative/integer` |
| gdbPercap             | `quantitative/numeric` |


`sapply` also gives each variable name and data type

```{r}
sapply(gapminder, class) 

```

## Exploring individual variables:

```{r}
head(gapminder) # this shows the first 6 rows and all columns of gapminder dataset.
attach(gapminder) #this makes it possible to use any variable of the gapminder dataset without using the `$` sign. i.e you can use gdpPercap instead of gapminder$gdpPercap.
```

### Exploring quantitative variable `gdpPercap`:

The `summary` command generates the minimum, 1st quartile, median, mean, 3rd quartile and maximum values of variable `gdpPercap`

```{r}
summary(gdpPercap) 
```

Using `ggplot` for a histogram to summarize the quantitative variable `gdpPercap` and by changing the number of bins to 50. This shows the spread/frequency of `gdpPercap`

```{r}
ggplot(gapminder, aes(x = gdpPercap)) + 
  geom_histogram(colour = "black", bins = 50, fill = "purple")

```


We can also make a kernel density plot of the variable `gdpPercap`. It gives a similar information as the histogram. It shows that gdpPercap around 10000 has the highest density.


```{r}
ggplot(gapminder, aes(gdpPercap)) + 
  geom_density(bw=0.1, fill = "forestgreen")

```

This is the combination of histogram and kernel density. It is possible to have the histogram and the kernel density on a single plot as below.

```{r}
ggplot(gapminder, aes(gdpPercap)) + 
  geom_histogram(aes(y=..density..), colour = "black", bins = 50, fill = "purple") +
  geom_density()

```


### Exploring categorical variable `continent`:
We want to know the number of countries in each continent by using the command `table`. The command `pie` shows the piechart which confirms that African continent has the highest number of countries and Oceania has the lowest. 

```{r}
Table <- table(continent)
Table  
pie(Table)

```

The command `barplot` shows the distribution of each continent.

```{r}
barplot(Table)

```
Making the piechart of the summary of variable `continent` gives a similar result as the `table` command.

```{r}
pie(summary(continent))

```

We can also use `ggplot` and `geom` to generate a similar barplot. It gives the frequency of each continent and also show that Africa has more countries while Oceania has less countries.

```{r}
ggplot(data = gapminder) +
  geom_bar(mapping = aes(x = continent, fill = continent))

```

Here I show how we can make a pie chart from a single categorical variable `continent` and the number countries in each continent using the command `coord_polar`. It shows how continent are partitioned. Here we use `theta=y`.

```{r}
b <- ggplot(data = gapminder) +
  geom_bar(mapping = aes(x = continent, fill = continent))
b + coord_polar(theta = "y")

```

This is the same as above, just that here, we use `theta=x`. 

```{r}
b <- ggplot(data = gapminder) +
  geom_bar(mapping = aes(x = continent, fill = continent))
b + coord_polar(theta = "x")

```
## Exploring various plot types:

### One quantitative and one categorical variable

#### Boxplots 
Here, I show the boxplot of gdpPercap among all continent using pink colour. I also use the `ggtitle` to specify the title of the plot. This shows that Oceania has the highest gdpPercap, while Africa has the lowest.


```{r}
ggplot(aes(x = continent, y = gdpPercap), data =gapminder) +
geom_boxplot(fill = "pink") +
ggtitle("gdpPercap among all continent")+theme(plot.title = element_text(hjust = 0.5))

```

This gives a similar result, but here we rescaled by using the command `scale_y_log10` to see a clearer difference between the boxplots.


```{r}
ggplot(aes(x = continent, y = gdpPercap), data =gapminder) +
  scale_y_log10() +
geom_boxplot(fill = "pink") +
ggtitle("gdpPercap among all continent")+theme(plot.title = element_text(hjust = 0.5))

```


Here, I show the boxplot of life expectancy in all continent using slategray1 colour. The command `scale_y_log10` is used for rescaling. This shows that Oceania has the highest lifeExp, while Africa has the lowest.

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  scale_y_log10() +
  geom_boxplot(fill = "slategray1")

```

#### Violin plots
This is a violin plot using the command `geom_violin`. It shows the distribution of life expectancy in each continent. We also see here that Oceania have higher life expectancy

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_violin(fill = "slategray1")

```


Similarly, this is a violin plot using the command `geom_violin` as well as rescaling to shows the distribution of gdpPercap in each continent. We also see here that Oceania have higher gdpPercap and Africa have the lowest.

```{r}
ggplot(gapminder, aes(continent, gdpPercap)) +
  scale_y_log10() +
  geom_violin(fill = "pink")

```

This is a violin plot with geometric point using the command `geom_violin` together with `geom_point`. It also shows the distribution of life expectancy in each continent. 

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_violin(fill = "slategray1") +
  geom_point(alpha=0.1)

```

It shows the distribution of gdpPercap in each continent. We also rescale here. 

```{r}
ggplot(gapminder, aes(continent, gdpPercap)) +
  scale_y_log10() +
  geom_violin(fill = "pink") +
  geom_point(alpha=0.1)

```


#### Jitter plots
This is a violin plot with jitter plot using the command `geom_jitter`. It also shows the distribution of gdpPercap in each continent. I rescaled here as well.

```{r}
ggplot(gapminder, aes(continent, gdpPercap)) +
  scale_y_log10() +
  geom_violin() +
  geom_jitter(alpha=0.3) 

```

This is also a violin plot with geometric point using the command `geom_violin` together with `geom_point`. It shows the distribution of life expectancy in each continent. 

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_violin() +
  geom_jitter(alpha=0.3)

```
This is the histogram of each continent versus their gdpPercap. The command `facet_wrap` is used to make separate histograms for each continent. The command `free_x` is used to make each continent have their own x-axisand not a shared x=axis. We can see here that Africa has the highest count of gdpPercap.

```{r}
ggplot(gapminder, aes(x = gdpPercap, fill = "colour")) + 
  geom_histogram(colour = 'black') + 
  facet_wrap('continent', scales = "free_x")

```  
  
### Two quantitative variables:

Scatterplot of `gdpPercap` vs. `lifeExp`. This is the scatter plot of gdpPercap and life expectancy using the command `geom_point`.

```{r}
ggplot(gapminder, aes(x=lifeExp, y=gdpPercap)) + geom_point()

```

This scatter plot of gdpPercap and life expectancy using the command `geom_line`.

```{r}
ggplot(gapminder, aes(lifeExp, gdpPercap)) +
  geom_line()

```

Here, we show the smooth curve of lifeExp versus gdpPercap using the command `geom_smooth`.

```{r}
ggplot(data=gapminder) +
  geom_smooth(mapping = aes(lifeExp, gdpPercap))

```
  

### Two quantitative and one categorical variables:

Here we write a code that:
    1. select variables country, lifeExp and gdpPercap from the data
    2. filters the data to Nigeria only
    3. pipes the filtered data into `ggplot`
    4. makes the time plot of `lifeExp` and `gdpPercap`
    5. displays the points
    6. add the line to join the points

```{r}
gapminder %>%
  select(country, lifeExp, gdpPercap) %>% 
  filter(country == "Nigeria") %>%
  ggplot(aes(lifeExp, gdpPercap)) +
  geom_line() +
  geom_point()
  
```

This is the scatter plot of `gdpPercap` versus `lifeExp` stratified by different continent. The command `facet_wrap` slpit the scatter plot into different continents. We can see from the plot below that the life expectancy of African continent settles below the gdpPercap of $30000$, while the life expectancy of Oceania continent between $71$ and $80$ settles have a gdpPercap between $10000$ to $40000$.


```{r}
ggplot(data=gapminder) +
  geom_point(mapping = aes(lifeExp, gdpPercap, color = continent)) + 
  facet_wrap(~ continent, scales = "free_x")

```
Using the command `facet_grid` and including `dot` with the command `facet_grid(.~ continent)` put all of the plots side by side one another for better comparison. This makes it obvious to see that African continent with life expectancy between $0$ and less than $80$ have their gdpPercap lower than $30000$.

```{r}
ggplot(data=gapminder) +
  geom_point(mapping = aes(lifeExp, gdpPercap, color = continent)) + 
  facet_grid(.~ continent)
  
```


The command `geom_smooth` produces a smooth curve with confidence interval and `colour=continent` splitted and coloured the curve by different continent. 

```{r}
ggplot(data=gapminder) +
  geom_smooth(mapping = aes(lifeExp, gdpPercap, color = continent))

```

This is the scatter plot of `gdpPercap` versus `lifeExp` coloured by different continent, The command `geom_point` gives different points of `gdpPercap` versus `lifeExp` of different continent. The command `geom_smooth` gives two lines that filters down to only African continent (the longer line) and Oceania continent (the shorter line).

```{r}
ggplot(data=gapminder, mapping = aes(lifeExp, gdpPercap)) +
  geom_point(mapping = aes(color = continent)) +
  geom_smooth(data = filter(gapminder, continent == "Africa")) +
  geom_smooth(data = filter(gapminder, continent == "Oceania"))

```

#### Relation between gdpPercap and lifeExp, adjusting for continent using boxplot.

We can categorize variable `lifeExp` so as to use it as a categorical variable. The plot shows the boxplot of gdpPercap versus categorize life expectancy and among different continent. The boxplots below show alot of information. Majority of Africans who have their life expectancy between $51-70$ have the highest gdpPercap. Asians have similar life expectancy across all group but with slightly different gdpPercap. Many Oceanians have their life expectancy between $41-50$ but with slightly different gdpPercap. 

```{r}
summary(lifeExp) #this give the minimum, 1st and 3rd quartile, median, mean and maximun.

Exp <- cut(lifeExp, breaks = c(0,40,50,70,85), label=c("<40","41-50", "51-70","71+")) #this line converts the quantitative variable lifeExp to a categorical variable of 4 levels.

ggplot(data = gapminder, aes(x = Exp, y = gdpPercap)) +
  scale_y_log10() +
  geom_boxplot(fill = "slategray1") +
  facet_wrap(~continent, scales = "free_x") 

```

#### Density plots of gdpPercap by continent from 1992 upward.
Here we filltered years above $1987$, pipes into ggplot and make a density plot of life expectancy of different continent in 4 different years. Across all years and relative to other continent, Oceania have the highest density of lifeExp.

```{r}

gapminder %>%
  filter(year > 1987) %>%
  ggplot(aes(x = lifeExp, fill = continent)) + facet_wrap(~year, scales = "free_x") +
  geom_density(alpha = 0.5)

```

#### Barplot of two quantitative and one categorical variable.

This is a multiple barplot comparing the gdpPercap of different continent versus their categorized life expectancy. I used `fill=continent` to have the barplot coloured by different continent and used `dodge` to have a separate barplot of different continent in each group and so as not to have them stacked on a barplot.

Here, for each group of life expectancy, we have all the continent. For example, among group with life expectancy $51-70$, the gdpPercap is highest in Asia . Similarly, among group with life expectancy $71+$, the gdpPercap is lowest in Africa.

```{r}
ggplot(gapminder, aes(fill=continent, y=gdpPercap, x=Exp)) + 
    geom_bar(position="dodge", stat="identity") +
  labs(x="Categorized life expectancy", y="gdpPercap")

```
This is a similar multiple barplot comparing the gdpPercap of different category of life expectancy among all continent. I used `fill=Exp` to have the barplot coloured by different category of lifeExp and similarly used `dodge` to have a separate barplot of different category of lifeExp in each continent and so as not to have them stacked on a barplot.

Here, for each continent, we have all group of life expectancy. For example in Africa, the group with life expectancy between $51-70$ have the highest gdpPercap. In Americas, the group with life expectancy around $71+$ have the highest gdpPercap. Similarly in Oceania, only two life expectancy group exist. i.e life expectancy of Oceanians falls between $51-71+$ and with $71+$ having the highest gdpPercap.


```{r}
ggplot(gapminder, aes(fill=Exp, y=gdpPercap, x=continent)) + 
    geom_bar(position="dodge", stat="identity") +
  labs(x="continent", y="gdpPercap")
```
This is a similar bar plot, just that `facet_wrap` was used to separate the barplot into different continent.

```{r}
ggplot(gapminder, aes(y=gdpPercap, x=Exp, fill=Exp)) + 
    geom_bar( stat="identity") +    
    facet_wrap(~continent, scales = "free_x")
```


### Three quantitative and one categorical variables:

Here we write a code that:
    1. select variables continent, year, lifeExp and gdpPercap from the data
    2. filters the data to year 2007 only
    3. pipes the filtered data into `ggplot`
    4. makes the time plot of `lifeExp` and `gdpPercap` and colour by continent
    5. displays the points
    6. add the title
    
We can see that most African countries with gdpPercap of $10000$ and lower have the their life expectancy between $30$ to $60$ in the year 2007. For all other continent, it is obvious that the most highest gdpPercap is around $10000$ with life expectancy between $70$ and $85$ in the same year.


    
```{r}
gapminder %>%
  select(continent, year, lifeExp, gdpPercap) %>% #select only continent, year, lifeExp, gdpPercap
  filter(year == "2007") %>% #filter down to only year 2007.
  ggplot(aes(lifeExp, gdpPercap, colour = continent )) +
  geom_point() +
  labs(title = "Scatterplot of gdpPercap in the year 2007 coloured by continent")

```

Here we write a code that:
    1. select variables continent, year, lifeExp and gdpPercap from the data
    2. filters the data to year 1952 only
    3. pipes the filtered data into `ggplot`
    4. makes the time plot of `lifeExp` and `gdpPercap` and colour by continent
    5. displays the points
    6. add the title

The interpretation of the graph is the same as above, just that this one is for the year 1952.

```{r}
gapminder %>%
  select(continent, year, lifeExp, gdpPercap) %>% #select only continent, year, lifeExp, gdpPercap
  filter(year == "1952") %>% #filter down to only year 2007.
  ggplot(aes(lifeExp, gdpPercap, colour = continent )) +
  geom_point() +
  labs(title = "Scatterplot of gdpPercap in the year 2007 coloured by continent")
```


### "Complete" plot

Plotting much of the data.

- lifeExp vs pop with gdpPercap bubbles
- facet by year
- colour by continent

```{r}
ggplot(gapminder, aes(lifeExp,pop)) +
  scale_y_log10() +
  geom_point(aes(size=gdpPercap, colour=continent)) +
  scale_size_area() +
  facet_wrap(~ year)
```

## I want to do more
Using either the command `%in%` or `|` gives what they wanted (24 rows).

```{r}
filter(gapminder, country ==c("Rwanda", "Afghanistan")) #produces 12 rows for variable country=="Rwanda" and country=="Afghanistan". This captures 6rows each and it is not what they want.

gapminder %>%
  filter(country =="Rwanda") #produces 12 rows for variable country=="Rwanda"


gapminder %>%
  filter(country =="Afghanistan") #produces 12 rows for variable country=="Afghanistan"

gapminder %>%
  filter(country %in% c("Rwanda", "Afghanistan")) #produces 24 rows for variable country=="Rwanda" and country=="Afghanistan". This is what they want.

gapminder %>%
  filter(country=="Rwanda" | country=="Afghanistan") #produces 24 rows for variable country=="Rwanda" and country=="Afghanistan". This is also what they want.

```

Here I want to use the `knitr::kable()` command to have a more fancy table. We will start by installing the necessary package and calling the libraries.

~~~~
install.packages("kableExtra")
~~~~

```{r}
library(knitr)
library(kableExtra)

```

Here I pipe and filter the country down to `Rwanda` and `Afghanistan`, use the `kable` command to have a fancy plot, use the command `kable_styling(full_width = F)` to shrink the size of the table.

Note: I decided not to display the kable code because it takes a longer time to run and makes the knitting slower. But here is the code:

~~~~
gapminder %>%
  filter(country=="Rwanda" | country=="Afghanistan") %>%
  kable() %>%
  kable_styling(full_width = F)
~~~~
